Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-01-02 article.qmd
Add Henning to authors

authors:
    -   name: Mauro Silberberg
        orcid: 0000-0002-2402-1100
        email: maurosilber@df.uba.ar
        affiliations:
        -   ref: UBA
        -   ref: CONICET
    -   name: Rahuman S. Malik-Sheriff
        orcid: 0000-0003-0705-9809
        email: sheriff@ebi.ac.uk
        affiliations:
        -   ref: EBI
    -   name: Hernán E. Grecco
        corresponding: true
        orcid: 0000-0002-1165-4320
        email: hgrecco@df.uba.ar
        affiliations:
        -   ref: UBA
        -   ref: CONICET

@@

```{python}
from simbio import Compartment, MassAction, Species, Reaction, initial

@@

    A: Species = initial(default=1)
    B: Species = initial(default=0)
    r = Reaction(
        reactants=[2 * A],
        products=[B],
        rate_law=1,
    )

@@

A reaction system for species $A$ and $B$ with initial conditions $1$ and $0$, respectively.
A single reaction transforming $2A$ into $B$ is saved in variable `r`.
The rate $1$ is specified directly for `Reaction`,
and is proportional to the reactants for `MassAction`.
::::

For the reaction networks,
our focus is on first-order differential equations
that describe the rate of change of species.
SimBio simplifies the definition of these network models by introducing `Species`,
which consists of a `poincare.Variable` and a stoichiometric number,
and `Reaction`,
a construct that converts reactants into products taking into account the stoichiometry (@fig-simbio).
Additionally, SimBio features `MassAction`, a subclass of `Reaction`,
which intuitively incorporates reactants into the rate law (@fig-simbio).

Several commonly used reactions are predefined as `MassAction` subclasses,
such as
`MichaelisMenten` ($S + E \leftrightarrow ES \rightarrow P + E$)
and its approximate form without the intermediate species $ES$,
and it is also simple to implement your own as subclasses of `Reaction` or `MassAction`.
Additionally, SimBio supports importing from and exporting to SBML,
and downloading them directly from BioModels [@malik-sheriffBioModels15Years2020] (@fig-simbio-io).

>>>>

authors:
    -   name: Mauro Silberberg
        orcid: 0000-0002-2402-1100
        email: maurosilber@df.uba.ar
        affiliations:
        -   ref: UBA
        -   ref: CONICET
    -   name: Rahuman S. Malik-Sheriff
        orcid: 0000-0003-0705-9809
        email: sheriff@ebi.ac.uk
        affiliations:
        -   ref: EBI
    -   name: Henning Hermjakob
        orcid: 0000-0001-8479-0262
        email: hhe@ebi.ac.uk
        affiliations:
        -   ref: EBI
    -   name: Hernán E. Grecco
        corresponding: true
        orcid: 0000-0002-1165-4320
        email: hgrecco@df.uba.ar
        affiliations:
        -   ref: UBA
        -   ref: CONICET

@@

```{python}
from simbio import Compartment, MassAction, Species, RateLaw, initial

@@

    A: Species = initial(default=1)
    B: Species = initial(default=0)
    r = RateLaw(
        reactants=[2 * A],
        products=[B],
        rate_law=1,
    )

@@

A reaction system for species $A$ and $B$ with initial conditions $1$ and $0$, respectively.
A single reaction transforming $2A$ into $B$ is saved in variable `r`.
The rate $1$ is specified directly for `RateLaw`,
and is proportional to the reactants for `MassAction`.
::::

For the reaction networks,
our focus is on first-order differential equations
that describe the rate of change of species.
SimBio simplifies the definition of these network models by introducing `Species`,
which consists of a `poincare.Variable` and a stoichiometric number,
and `RateLaw`,
a construct that converts reactants into products taking into account the stoichiometry (@fig-simbio).
Additionally, SimBio features `MassAction`, a subclass of `RateLaw`,
which intuitively incorporates reactants into the rate law (@fig-simbio).

Several commonly used reactions are predefined as `MassAction` subclasses,
such as
`MichaelisMenten` ($S + E \leftrightarrow ES \rightarrow P + E$)
and its approximate form without the intermediate species $ES$,
and it is also simple to implement your own as subclasses of `RateLaw` or `MassAction`.
Additionally, SimBio supports importing from and exporting to SBML,
and downloading them directly from BioModels [@malik-sheriffBioModels15Years2020] (@fig-simbio-io).
